---
title: "Undergrad Biostatistics - R Training - Week V"
author: "Ege Ulgen"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# One-sample t Test

We will use the AIDS data again (this time reading from a URL:

```{r read}
URL <- "https://raw.githubusercontent.com/egeulgen/MED131_22_23/main/data/aids_dataset.txt"
aids_df <- read.delim(URL, sep = " ")
```

Let's explore the data:
```{r eda}
# print first 6 rows (default)
head(aids_df)

# how many patients are there in each treatment group?
table(aids_df$treatment)

# summary of all variables
summary(aids_df)

boxplot(aids_df$cd4_2)
boxplot(aids_df$cd4_2~aids_df$treatment)
```

We'll use the "trt1" subset and perform hypothesis tests for CD4 at second measurement:

```{r trt1}
# subset for trt1
trt1_df <- subset(aids_df, treatment == "trt1")

summary(trt1_df$cd4_2)
hist(trt1_df$cd4_2)
```


```{r}
?t.test

# is the CD4 level at week 2 for trt1 different than 0?
t.test(trt1_df$cd4_2)

# is the CD4 level at week 2 for trt1 different than 20?
t.test(trt1_df$cd4_2, mu = 20)

# is the CD4 level at week 2 for trt1 different than 25?
t.test(trt1_df$cd4_2, mu = 25)


# is the CD4 level at week 2 for trt1 different than 20? (alpha = 0.1, conf. level = 90%)
t.test(trt1_df$cd4_2, mu = 20, conf.level = 0.9)

# is the CD4 level at week 2 for trt1 different than 20? (alpha = 0.01, conf. level = 99%)
t.test(trt1_df$cd4_2, mu = 20, conf.level = 0.99)


# is the CD4 level at week 2 for trt1 larger than 25?
t.test(trt1_df$cd4_2, mu = 25, alternative = "greater")

# is the CD4 level at week 2 for trt1 smaller than 60?
t.test(trt1_df$cd4_2, mu = 60, alternative = "less")
```

# Two-sample t Test

We'll use the `genderweight` data from the `datarium` package for this exercise. First, let's install the package (you'll need to this only once):

```{r pkg}
# install.packages("datarium")
```

Let's load and explore the data:

```{r}
data("genderweight", package = "datarium")

head(genderweight)

# frequency table of gender (group)
table(genderweight$group)

# histogram of weight (overall)
hist(genderweight$weight)

# boxplot of weight (overall)
boxplot(genderweight$weight)

# boxplot of weight by gender
boxplot(genderweight$weight~genderweight$group)

# mean weight of men
mean(genderweight$weight[genderweight$group == "M"])

# mean weight of women
mean(genderweight$weight[genderweight$group == "F"])

# sd of weight of men
sd(genderweight$weight[genderweight$group == "M"])

# sd of weight of women
sd(genderweight$weight[genderweight$group == "F"])
```

Is the mean weight of men significantly different than the mean weight of women?

```{r twosamptest}
t.test(weight~group, data = genderweight)
```

